Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-24 paper.tex
Cites for DP/GQ

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices~\citep[e.g.][]{chen2024genomic}.
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
genotype missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks require negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
The transition from the manifold inefficiencies of 
present-day ``copy-oriented'' computing,
to the ``mask-oriented'' analysis of large immutable, single-source
datasets is a potentially transformational change enabled by Zarr.

>>>>

Finally, to illustrate performance on a common filtering task, we 
created a copy of the VCF chunk which contains only variants 
that pass some common filtering criteria using
\texttt{bcftools view -I --include "FORMAT/DP>10 \& FORMAT/GQ>20"},
following standard practices~\citep[e.g.][]{bergstrom2020insights,
kousathanas2022whole,
chen2024genomic}.
This took a total of 689 min CPU time, with the same elapsed time.
In comparison, computing and storing a variant mask (i.e., a boolean value 
for each variant denoting whether it should be considered or not
for analysis) based on the same criteria using Zarr 
required a total of 1.96 min CPU time with an elapsed time 
of  11 min (63X faster).
%689 / 11 = 62.6 = 63X
There is an important distinction here between creating a copy
of the data (an implicit part of VCF based workflows) and creating an
additional \emph{mask}. As Table~\ref{tab-genomics-england-data} 
illustrates, call-level masks are cheap (the standard 
genotype missingness mask, call\_genotype\_mask, uses 0.49\% of the overall 
storage) and variant or sample level masks require negligible storage.
If downstream software can 
use configurable masks (at variant, sample and call level) 
rather than expecting full copies of the data, major storage savings
and improvements in processing efficiency can be made.
The transition from the manifold inefficiencies of 
present-day ``copy-oriented'' computing,
to the ``mask-oriented'' analysis of large immutable, single-source
datasets is a potentially transformational change enabled by Zarr.
